---
title: "Comparing MDAT diversity metrics"
author: "emily shumchenia"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: github_document
---

```{r load data, include=FALSE}

## required packages
library(dplyr)
library(raster)
library(viridis)
library(Hmisc)
library(knitr)
library(kableExtra)
library(rPython)

## load data: these are rasters of species richness, shannon index, and simpson index for mdat cetacean models.

cet_richness <-raster("~/Dropbox/Research/mdat_diversity_indices/data/mammal_richness_Cetaceans.tif")
cet_simpson <-raster("~/Dropbox/Research/mdat_diversity_indices/data/mammal_simpson_Cetaceans.tif")
cet_shannon <-raster("~/Dropbox/Research/mdat_diversity_indices/data/mammal_shannon_Cetaceans.tif")

## clip all rasters to the same extent as richness
cet_simpson <-crop(cet_simpson, extent(-347000, 1803000, -1081000, 1519000))
cet_shannon <-crop(cet_shannon, extent(-347000, 1803000, -1081000, 1519000))

```
## Take-home message
* If you are concerned about accounting for and considering rare species, **SPECIES RICHNESS** does the best job, by representing every species equally, regardless of abundance
* If you care about rare species, but need to give some additional consideration to species that are particularly abundant, the **SHANNON INDEX** does a good job
* If abundant or dominant species are important for your analysis, the **SIMPSON INDEX** gives more weight to highly abundant species, to the detriment of the representation of the rarest species

## Introduction
What are diversity metrics?  

When most people think about diversity, they think about the total number of species in an area. However, the total number of species is actually a metric called Species Richness.  

Diversity indices take the total number of species *and* their abundances into account (in different ways, depending on the index).  

Because each metric integrates information differently, it is important to know how they are similar or different, and how the outputs could be used.


##Purpose

MDAT has represented diversity with three different metrics: Species Richness, the Shannon Index, and the Simpson Index. The purpose of this document is to explore a few questions about these metrics:  

* How are the metrics similar or different?
* How do their  results compare to each other?
* Do they each capture similar ecological patterns?

Here are maps (top row) and histograms (bottom row) for Species Richness, Shannon Diversity, and Simpson Diversity for all cetacean species:  

```{r maps, echo=FALSE}

## plot the three maps
par(mfrow=c(2,3))
plot(cet_richness, main="Cetacean Species Richness", col=viridis(256), axes=FALSE)
plot(cet_shannon, main="Cetacean Shannon Diversity", col=viridis(256), axes=FALSE)
plot(cet_simpson, main="Cetacean Simpson Diversity", col=viridis(256), axes=FALSE)
hist(cet_richness, main=NULL, col=viridis(12), xlab="Species Richness")
hist(cet_shannon, main=NULL,  col=viridis(13), xlab="Shannon Diversity")
hist(cet_simpson, main=NULL,  col=viridis(18), xlab="Simpson Diversity")

```

##Quick comparison
From the maps and histograms we can tell that Shannon and Simpson are more similar to each other than either of them are to Species Richness. We can check this by doing a simple correlation between each pair of metrics:  

```{r metrics_corr, echo=FALSE, results='asis'}

## make a raster stack for each species group's set of diversity metrics
cet_stack <- stack(cet_richness, cet_simpson, cet_shannon)

## extract each stack's values to a table
cet_metrics <-values(cet_stack)

## change the column headers
colnames(cet_metrics) <- c("richness", "simpson", "shannon")

## calculate correlations among diversity metrics, save as data frame object, display the table
cet_metrics_corr <-rcorr(as.matrix(cet_metrics))
df_metrics_corr <-data.frame(cet_metrics_corr$r)
kable(df_metrics_corr, digits=2)

```

This result is not surprising since both Shannon and Simpson indices consider abundance information, and Species Richness does not.

##Comparing numeric results
To delve more deeply into how each index deals with abundance information, we can plot the Shannon and Simpson results together, showing how abundance information at each location has contributed to each diversity index value for that location:  

[insert plot from Jesse]  

This plot shows that for low to medium abundance values, both the Shannon Index and Simpson index generate comparable diversity results. However, for high abundance values, the Simpson index generates a proportionally much higher diversity value than the Shannon index.  

This result indicates that the Simpson index gives more weight to highly abundant species than the Shannon index. The Shannon index treats rare species and dominant species more equally.

##Ecological relationships
```{r regressions, include=FALSE}

## load bathymetry data and crop to the same extent as richness
bathy <-raster("~/Dropbox/Research/mdat_diversity_indices/data/ETOPO1_mdat.tif")
bathy <-crop(bathy, extent(-347000, 1803000, -1081000, 1519000))

## make the cet_metrics object into a dataframe, omitting NAs
cet_metrics_df <-data.frame(cet_metrics)
cet_metrics_df <-na.omit(cet_metrics_df)

## extract water depth values from bathy raster to a data object, omitting NAs
depths <-values(bathy)
depths <-data.frame(depths)
depths <-na.omit(depths)

## calculate regressions of metrics and water depth
richness_depth <-lm(cet_metrics_df$richness ~ depths$depths)
summary(richness_depth)

shannon_depth <-lm(cet_metrics_df$shannon ~ depths$depths)
summary(shannon_depth)

simpson_depth <-lm(cet_metrics_df$simpson ~ depths$depths)
summary(simpson_depth)

## make table of the results
results <-matrix(c(.41, .001, .29, .001, .19, .001), ncol=2, byrow=TRUE)
colnames(results) <- c("r", "p")
rownames(results) <- c("richness", "shannon", "simpson")
results <-as.table(results)

```
Does each metric represent diversity similarly across different ocean habitats? In other words, will each metric give us the same answer to the question: "which habitats have high diversity and which have lower diversity?"   

An important cetacean habitat variable is depth. A linear regression of each diveristy metric on water depth will help explain whether the three metrics respond similarly to real changes in diversity that may occur with depth.  

```{r regresion results, echo=FALSE, results='asis'}

par(mfrow=c(1,2))
plot(bathy, main="Water depth (m)", col=magma(256), axes=FALSE)
kable(results, caption="Linear regression with depth", format="html") %>%
  kable_styling(bootstrap_options = "striped", full_width=FALSE)

```
